\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-29 tex/paper.tex
<:(((>< example

\glnote{contour plots of q(theta|psi) + histograms + learning curves}
\glnote{contours maybe with $\gamma=0$ and $\gamma>0$ contours}
\glnote{insist that we dont learn the simulator}

As a first illustrative experiment, we evaluate adversarial variational
optimization on a Poisson distribution with unknown scale parameter $\lambda$.
We artificially consider the distribution as a parameterized black box
simulator, from which we can only generate (discrete) data.

The observed data is sampled from a Poisson with scale $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 300 epochs with mini-batches of size $M=64$
and the following configuration. For the critic, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=100$ iterations with a step size $\alpha=0.01$, decay rates
$\beta_1=\beta_2=0.5$ and its inner first and second moment vectors reset at
each epoch in order to avoid building momentum in staled directions.  For
estimating $\lambda^*$, we use a univariate Gaussian proposal distribution
$q(\log \lambda|\psi)$ initialized with a mean at $\log(5)$ and unit variance. At
each epoch, parameters $\psi$ are updated by making one Adam step, with
$\alpha=0.01$ and $\beta_1=\beta_2=0.1$. Penalty coefficients are set to
$\lambda=0.025$ and $\gamma=5$.

\begin{figure}
    \centering
    \vspace{-0.5cm}
    \includegraphics[width=0.48\textwidth]{figures/poisson-gamma=5.pdf}
    \vspace{-1cm}
    \caption{Toy problem on discrete data: Poisson with unknown scale.
             This example illustrates adversarial variational optimization for a discrete non-differentiable generator.
             (Top) Model distribution $p(\mathbf{x}|\psi)$ after adversarial variational optimization along with the proposal distribution $q(\lambda|\psi)$.
             (Bottom) Evolution of the variational upper bound $U_d$ on the loss ${\cal L}_d$ of the critic.
             }\label{fig:poisson}
\end{figure}

The upper plot of Figure~\ref{fig:poisson} illustrates the resulting model
after the optimization procedure. As theoretically expected from adversarial
training, the model density $p(\mathbf{x}|\psi)$ closely matches the true
density $p_r(\mathbf{x})$, as shown by the aligned histograms in blue and
orange. Under the effect of the entropy penalty $\gamma H(q(\theta|\psi))$, we
also observe that the proposal distribution has correctly concentrated its
mass around $\log(7) = 1.94$. The bottom plot of
Figure~\ref{fig:poisson} shows an empirical estimate of $-U_d$ with respect to
the epoch. We observe that it quickly falls towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)}
[d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$ and that the critic cannot distinguish
between true and model data. Despite the discreteness and the non-differentiability
of the underlying generator, this confirms that inference with
adversarial variational optimization works.

@@

\glnote{use 5d example from carl}
\glnote{contour plots}

>>>>

As a first illustrative experiment, we evaluate inference for a discrete Poisson
distribution with unknown scale parameter $\lambda$. We artificially consider
the distribution as a parameterized simulator, from which we can only
generate data.

The observed data is sampled from a Poisson with scale $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 300 epochs with mini-batches of size $M=64$
and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=100$ iterations with a step size $\alpha=0.01$, decay rates
$\beta_1=\beta_2=0.5$ and its inner first and second moment vectors reset at
each outer epoch in order to avoid building momentum in staled directions.  For
estimating $\lambda^*$, we use a univariate Gaussian proposal distribution
$q(\log \lambda|\psi)$ initialized with a mean at $\log(5)$ and unit variance. At
each epoch, parameters $\psi$ are updated by taking one Adam step, with
$\alpha=0.01$ and $\beta_1=\beta_2=0.1$. The gradient penalty coefficient is set to
$\lambda=0.025$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
    \caption{Discrete Poisson model with unknown scale.
             (Top left) Proposal distributions $p(\theta|\psi)$ after adversarial variational optimization. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density and therefore to better inference.
             (Top right) Model distributions $p(\mathbf{x}|\psi)$ after training.
             (Bottom) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:poisson}
\end{figure}

The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
distributions $p(\theta|\psi)$ after adversarial variational optimization.  For
both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
the effect of the positive entropy penalty $H(q(\theta|\psi))$,
the proposal distribution for $\gamma=5$ concentrates its mass more tightly,
yielding in this case a more precise inference.  The top right plot compares the
model distributions to the true distribution.  As theoretically expected from
adversarial training, we see that the resulting distributions closely match
the true distribution, with qualitatively  better results for the penalized
model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=5$,
the curves quickly fall towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)}
[d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$ and that the critic cannot distinguish
between true and model data. Despite the discreteness and the
non-differentiability of the underlying generator, this confirms that inference
with adversarial variational optimization works.

@@

\glnote{use 5d example from carl.}
\glnote{contour plots for $\alpha$ and $\beta$.}
